Synthesizing Symmetric Lenses

### 2018-07-10 papers/SymmetricLenses/main.tex
add response to steve comments

The expected number of bits required to recover data is a well-known concept in
information theory: it is the \emph{entropy} of the data. \saz{Cite Shannon?} Calculating the
entropy of data requires a probability distribution over the space of possible
values for the data. Specifically, given a set $S$ and a probability
distribution over that set $P \OfType S \rightarrow \Reals$, the entropy of $S$
with probability distribution $P$ is $-\Sigma_{s\in S}P(s)\Log_2P(s)$.

@@

\section{Simple Symmetric String Lenses}
\label{sec:ssl}
\saz{This paragraph used to say that lenses are typed between stochastic REs
  and it also says that lenses are typed between REs.  I think its should just
  be REs? Double check that this is correct}

We now give a formal description of all the combinators in our symmetric lens
language.  While our algorithms perform synthesis on stochastic regular
expressions, lenses themselves are typed between regular expressions.  We can
recover a regular expression from a stochastic regular expression by deleting
probability annotations.  In a slight abuse of notation\saz{Why is this an
  abuse?}, if $\Regex$ is a stochastic regular expression, $\BRegex$ is the
regular expression obtained by removing probability annotations. The typing
judgment $\Lens \OfType \BRegex \Leftrightarrow \BRegexAlt$ means that $\Lens$
is a well typed lens between $\LanguageOf{\BRegex}$ and
$\LanguageOf{\BRegexAlt}$.

@@

% begin synthesis
\section{Synthesis}
\label{sec:synthesis}
Algorithm~\ref{alg:synth-sym-lens} presents our synthesis algorithm at a high
level of abstraction. The input regular expressions are first converted into
stochastic regular expressions with \ToStochastic.  
%Once that conversion has
%taken place, the $\RXSearch$ algorithm executes. This algorithm itself is
%defined via five functions: (1) $\PCF{\PQ.Create}$, (2) $\PCF{\PQ.Pop}$, (3)
%$\PCF{Expand}$, (4) $\PCF{\PQ.Push}$, and (5) $\PCF{\Continue}$. First,
This pair of SREs is used to initialize a priority queue ($pq$).
Next, $\PCF{SynthSymLens}$
enters a loop that searches for likely lenses. The loop terminates when the
algorithm believes it is unlikely to find a better lens than the best one it has
found so far (a termination condition defined by $\PCF{\Continue}$). Within each
iteration of the loop, it:
\begin{itemize}
\item pops the next class ($S$, $T$) of lenses to
search off of the priority queue ($\PCF{\PQ.Pop}$),
\item executes $\GreedySynth$ to find a best lens in that class if
one exists ($\LC$),
using the examples $exs$ to filter out potential lenses that do not satisfy
the specification,
\item replaces $best$ with $\LC$, if $\LC$ has lower cost according
to our information theoretic metric, and
\item adds all of the SREs derived from rewriting $S$ and $T$ ($\Expand(S,T)$)
  to the priority queue with $\PCF{\PQ.Push}$ \saz{what determines the priority order?}
\end{itemize}
When the loop terminates, the search returns the globally best lens found
($best$).  Each element of this algorithm will be explained in
further depth in the following subsections.

@@

\subsection{Search Through Types}
The first synthesizer is one that searches through regular expression pairs.
This synthesizer, comprising \Expand, \PQ, and \RXSearch \saz{The name \RXSearch
  does not appear in the figure.  I get the impression that it is the name of
  the while loop, but perhaps we can make that clearer in the algorithm code?
  Maybe by putting a brace around the while loop and giving it a name? Here we
  could say something like ``... and a loop that continues searching until a
  satisfactory lens is found (we call this loop \RXSearch)}.  It proposes regular
expressions that are semantically, but not syntactically, equivalent to the originals. In
particular, this synthesizer performs the following rewrites:
\begin{center}
  \begin{tabular}{rcl}
    $\PRegexStar{\Regex}{\Probability}$
    & \Rewrite
    & \PRegexOr{\EmptyString{}}{(\RegexConcat{\Regex{}}{\PRegexStar{\Regex{}}{\Probability}})}{\Probability}\\

@@

\subsection{\GreedySynth}
\label{subsec:greedy-synth}
\saz{We should call out the structural similarity
  between the DNF grammar and the organization of these algorithms.}
The greedy algorithm is actually two\saz{three! don't forget AtomSynth} greedy algorithms, one that greedily finds
symmetric DNF lenses (\GreedySynth) and one that greedily finds symmetric
sequence lenses (\GreedySeqSynth). We describe this synthesis algorithm in the
context of trying to find a lens between \lstinline{"" | name.name*} and
\lstinline{"" | name}.

@@

First, \PCF{CannotMap} checks whether there is no possible lens satisfying the
examples. If so, \GreedySynth returns \None early. Otherwise, \GreedySynth finds
the best lenses (given the examples that match them) between all sequence pairs
$(SQ_i, TQ_j)$ drawn from the left and right DNF regular expressions. (The
function \CartesianMap maps its argument across the cross product of the input
lists).  A priority queue containing these sequence lenses is then initialized
with $\PCF{PQ.Create}$.\saz{Ordered how?} The symmetric lens is then built up
iteratively from these sequence lenses, where the state of the partially constructed lens is
tracked in the lens builder, $lb$.

@@

Finally, after all sequences have been popped, the partial DNF lens $lb$ is
converted into a symmetric DNF lens. This is only possible if all sequences are
involved in some sequence lens: if they are not, \PCF{GreedyState.ToDNFLens}
instead returns \None.

\begin{algorithm}
  \caption{\GreedySynth}
  \label{alg:greedysynth}
  \begin{algorithmic}[1]
    \Function{GreedySynth}{$\Examples,\DNFOf{(\Sequence_1,\Probability_1)\DNFSep\ldots\DNFSep(\Sequence_n,\Probability_n)},\DNFOf{(\SequenceAlt_1,\ProbabilityAlt_1)\DNFSep\ldots\DNFSep(\SequenceAlt_m,\ProbabilityAlt_m)}$}
    \If{$\Call{CannotMap}{\Examples,\DNFOf{(\Sequence_1,\Probability_1)\DNFSep\ldots\DNFSep(\Sequence_n,\Probability_n)},\DNFOf{(\SequenceAlt_1,\ProbabilityAlt_1)\DNFSep\ldots\DNFSep(\SequenceAlt_m,\ProbabilityAlt_m)}}$}
    \State $\ReturnVal{\None}$
    \EndIf
    \State $\SLS \gets
    \Call{\CartesianMap}{\GreedySeqSynth(\Examples),\ListOf{\Sequence_1;\ldots;\Sequence_n},\ListOf{\SequenceAlt_1;\ldots;\SequenceAlt_m}}$
    \State $\GreedyState \gets \Call{PQ.Create}{\SLS}$
    \State $lb \gets \Call{LensBuilder.Empty}{}$
    \While{$\Call{PQ.IsNonempty}{\GreedyState}$}
    \State $\SSQLens \gets
    \Call{PQ.Pop}{\GreedyState}$
    \If{$\Call{LensBuilder.UsefulAdd}{lb,\SSQLens,\Examples}$}
    \State $lb \gets \Call{GreedyState.AddSeq}{lb,\SSQLens}$
    \EndIf
    \EndWhile
    \State \ReturnVal{\Call{LensBuilder.ToDNFLens}{\GreedyState}}
    \EndFunction
  \end{algorithmic}
\end{algorithm}

@@

The inputs to \GreedySeqSynth are a suite of input-output
examples and a pair of stochastic atoms. As in \GreedySynth, \GreedySeqSynth
returns \None early if there is no possible lens. Afterward,
\GreedySeqSynth finds the best lenses between each atom pair of the left and
right sequences, and organizes them into a priority queue 
with \PCF{PQ.Create}. The symmetric sequence lens is built up iteratively from
these atom lenses, where the state of the partially built lens is tracked in the sequence lens
builder, $slb$.

\GreedySeqSynth then loops until there are no more atom lenses in the priority
queue. Within this loop, atom lenses are popped from the queue and the
``useful'' ones are kept.  An atom lens is useful if adding it will lower the cost of the
generated sequence lens, or if examples show that one of the atoms must be
connected to another lens.\saz{I'm not sure what ``must be connected to another
  lens means''}  The priorities of these lenses change over time, because
each atom can only be involved in one lens. Adding a lens may necessitate
removing another, so not all popped atom lenses will lower the cost of the
generated sequence lens.\saz{I don't understand the last two sentences.}

Finally, after all atoms have been popped, the sequence lens is validated and
returned. The algorithm confirms that all atoms that need to be connected with
an atom lens, are. If they are not, \PCF{SLensBuilder.ToDNFLens} instead returns
\None.

@@

Our implementation includes a number of optimizations to this algorithm.  We added ``open'' and
``closed'' annotations to regular expressions \saz{what do these mean? I guess
  ``open'' means ``subject to  DNF conversion'' and ``closed'' means not?}, an expansion inference algorithm,
hash-consing and memoization, along with compositional synthesis. The implementation
also allows an additional parameter in specifications that customizes the
termination condition for the synthesis procedure.

\paragraph*{Open and Closed Regular Expressions} While \GreedySynth acts
relatively efficiently, it suffers from an exponential blowup when converting
regular expressions to DNF form. This can be minimized by not converting
\emph{all} of the regular expression into DNF form, but only converting
\emph{some} of it to DNF form, and doing so incrementally.  This change is easy
to make by adding non-DNF stochastic regular expressions as a new type of
``closed'' atom.  When converted into a DNF regular expression, if $\Regex$ is
such a closed stochastic regular expression, then
$\ToDNFRegexOf{\Regex} = \AtomToDNFOf{\Regex}$.

@@

Classical symmetric lenses satisfy the following equational laws.
\begin{equation}
  \tag{\PutRL}
  \frac{\PutRSymOf{(x,c)} = (y,c'')}{\PutLSymOf{(y,c'''')} = (x,c')}
\end{equation}
\begin{equation}
  \tag{\PutLR}
  \frac{\PutLSymOf{(y,c)} = (x,c')}{\PutRSymOf{(x,c')} = (y,c')}
\end{equation}

>>>>

The expected number of bits required to recover data is a well-known concept in
information theory: it is the \emph{entropy} of the data~\cite{Shannon1948}.
Calculating the entropy of data requires a probability distribution over the
space of possible values for the data. Specifically, given a set $S$ and a
probability distribution over that set $P \OfType S \rightarrow \Reals$, the
entropy of $S$ with probability distribution $P$ is $-\Sigma_{s\in
  S}P(s)\Log_2P(s)$.

@@

\section{Simple Symmetric String Lenses}
\label{sec:ssl}
We now give a formal description of all the combinators in our symmetric lens
language. While our algorithms perform synthesis on stochastic regular
expressions, lenses themselves are typed between regular expressions. We can
recover a regular expression from a stochastic regular expression by deleting
probability annotations. In a slight abuse of notation\saz{Why is this an
  abuse?}\afm{We also use \BRegex{},\BRegexAlt{} as nonterminals}, if $\Regex$
is a stochastic regular expression, $\BRegex$ is the regular expression obtained
by removing probability annotations. The typing judgment $\Lens \OfType \BRegex
\Leftrightarrow \BRegexAlt$ means that $\Lens$ is a well typed lens between
$\LanguageOf{\BRegex}$ and $\LanguageOf{\BRegexAlt}$.

@@

% begin synthesis
\section{Synthesis}
\label{sec:synthesis}
Algorithm~\ref{alg:synth-sym-lens} presents our synthesis algorithm at a high
level of abstraction. The input regular expressions are first converted into
stochastic regular expressions with \ToStochastic.  
%Once that conversion has
%taken place, the $\RXSearch$ algorithm executes. This algorithm itself is
%defined via five functions: (1) $\PCF{\PQ.Create}$, (2) $\PCF{\PQ.Pop}$, (3)
%$\PCF{Expand}$, (4) $\PCF{\PQ.Push}$, and (5) $\PCF{\Continue}$. First,
This pair of SREs is used to initialize a priority queue ($pq$).
Next, $\PCF{SynthSymLens}$
enters a loop that searches for likely lenses. The loop terminates when the
algorithm believes it is unlikely to find a better lens than the best one it has
found so far (a termination condition defined by $\PCF{\Continue}$). Within each
iteration of the loop, it:
\begin{itemize}
\item pops the next class ($S$, $T$) of lenses to
search off of the priority queue ($\PCF{\PQ.Pop}$),
\item executes $\GreedySynth$ to find a best lens in that class if
one exists ($\LC$),
using the examples $exs$ to filter out potential lenses that do not satisfy
the specification,
\item replaces $best$ with $\LC$, if $\LC$ has lower cost according
to our information theoretic metric, and
\item adds all of the SREs derived from rewriting $S$ and $T$ ($\Expand(S,T)$)
  to the priority queue with $\PCF{\PQ.Push}$. The priority of a SRE pair is the
  number of rewrites needed to derive the pair from the originals.
\end{itemize}
When the loop terminates, the search returns the globally best lens found
($best$).  Each element of this algorithm will be explained in
further depth in the following subsections.

@@

\subsection{Search Through Types}
The first synthesizer is one that searches through regular expression pairs.
This synthesizer comprises \Expand, \PQ, and \Continue. It proposes regular
expressions that are semantically, but not syntactically, equivalent to the
originals. In particular, this synthesizer performs the following rewrites:
\begin{center}
  \begin{tabular}{rcl}
    $\PRegexStar{\Regex}{\Probability}$
    & \Rewrite
    & \PRegexOr{\EmptyString{}}{(\RegexConcat{\Regex{}}{\PRegexStar{\Regex{}}{\Probability}})}{\Probability}\\

@@

\subsection{\GreedySynth}
\label{subsec:greedy-synth}
The greedy algorithm is actually three algorithms, one that greedily finds
symmetric DNF lenses (\GreedySynth), one that greedily finds symmetric sequence
lenses (\GreedySeqSynth), and one that finds symmetric atom lenses
(\AtomSynth{}). These three algorithms are hierarchically structured:
\GreedySynth relies on \GreedySeqSynth, \GreedySeqSynth relies on \AtomSynth,
and \AtomSynth rlies on \GreedySynth. The structure of the algorithms mirrors
the structure of symmetric DNF lenses and \SDNFREabbrevs. We describe this
synthesis algorithm in the context of trying to find a lens between
\lstinline{"" | name.name*} and \lstinline{"" | name}.

@@

First, \PCF{CannotMap} checks whether there is no possible lens satisfying the
examples. If so, \GreedySynth returns \None early. Otherwise, \GreedySynth finds
the best lenses (given the examples that match them) between all sequence pairs
$(SQ_i, TQ_j)$ drawn from the left and right DNF regular expressions. (The
function \CartesianMap maps its argument across the cross product of the input
lists). A priority queue containing these sequence lenses, ordered by cost, is
then initialized with $\PCF{PQ.Create}$. The symmetric lens is then built up
iteratively from these sequence lenses, where the state of the partially
constructed lens is tracked in the lens builder, $lb$.

@@

Finally, after all sequences have been popped, the partial DNF lens $lb$ is
converted into a symmetric DNF lens. This is only possible if all sequences are
involved in some sequence lens: if they are not, \PCF{LensBuilder.ToDNFLens}
instead returns \None.

\begin{algorithm}
  \caption{\GreedySynth}
  \label{alg:greedysynth}
  \begin{algorithmic}[1]
    \Function{GreedySynth}{$\Examples,\DNFOf{(\Sequence_1,\Probability_1)\DNFSep\ldots\DNFSep(\Sequence_n,\Probability_n)},\DNFOf{(\SequenceAlt_1,\ProbabilityAlt_1)\DNFSep\ldots\DNFSep(\SequenceAlt_m,\ProbabilityAlt_m)}$}
    \If{$\Call{CannotMap}{\Examples,\DNFOf{(\Sequence_1,\Probability_1)\DNFSep\ldots\DNFSep(\Sequence_n,\Probability_n)},\DNFOf{(\SequenceAlt_1,\ProbabilityAlt_1)\DNFSep\ldots\DNFSep(\SequenceAlt_m,\ProbabilityAlt_m)}}$}
    \State $\ReturnVal{\None}$
    \EndIf
    \State $\SLS \gets
    \Call{\CartesianMap}{\GreedySeqSynth(\Examples),\ListOf{\Sequence_1;\ldots;\Sequence_n},\ListOf{\SequenceAlt_1;\ldots;\SequenceAlt_m}}$
    \State $\GreedyState \gets \Call{PQ.Create}{\SLS}$
    \State $lb \gets \Call{LensBuilder.Empty}{}$
    \While{$\Call{PQ.IsNonempty}{\GreedyState}$}
    \State $\SSQLens \gets
    \Call{PQ.Pop}{\GreedyState}$
    \If{$\Call{LensBuilder.UsefulAdd}{lb,\SSQLens,\Examples}$}
    \State $lb \gets \Call{LensBuilder.AddSeq}{lb,\SSQLens}$
    \EndIf
    \EndWhile
    \State \ReturnVal{\Call{LensBuilder.ToDNFLens}{\GreedyState}}
    \EndFunction
  \end{algorithmic}
\end{algorithm}

@@

The inputs to \GreedySeqSynth are a suite of input-output examples and a pair of
stochastic atoms. As in \GreedySynth, \GreedySeqSynth returns \None early if
there is no possible lens. Afterward, \GreedySeqSynth finds the best lenses
between each atom pair of the left and right sequences, and organizes them into
a priority queue ordered by cost with \PCF{PQ.Create}. The symmetric sequence
lens is built up iteratively from these atom lenses, where the state of the
partially built lens is tracked in the sequence lens builder, $slb$.

\GreedySeqSynth then loops until there are no more atom lenses in the priority
queue. Within this loop, atom lenses are popped from the queue and the
``useful'' ones are kept. An atom lens is useful if adding it will lower the
cost of the generated sequence lens, or if examples show that one of the atoms
must not be disconnected. Each atom can only be part of one lens at a time, so
the algorithm must at times disconnect previously connected atoms to connect an
atom that must not be disconnected.

Finally, after all atoms have been popped, the sequence lens is validated and
returned. The algorithm confirms that all atoms that must not be disconnected
are involved in an atom lens. If they are not, \PCF{SLensBuilder.ToDNFLens}
instead returns \None.

@@

Our implementation includes a number of optimizations to this algorithm. We
added ``closed'' annotations to regular expressions \saz{what do these mean? I
  guess ``open'' means ``subject to DNF conversion'' and ``closed'' means
  not?}\afm{I describe in greater detail below, should I add in additional info
  here too?}, an expansion inference algorithm, hash-consing and memoization,
along with compositional synthesis. The implementation also allows an additional
parameter in specifications that customizes the termination condition for the
synthesis procedure.

\paragraph*{Open and Closed Regular Expressions} While \GreedySynth acts
relatively efficiently, it suffers from an exponential blowup when converting
regular expressions to DNF form. This can be minimized by not converting
\emph{all} of the regular expression into DNF form, but only converting
\emph{some} of it to DNF form, and doing so incrementally. This change is easy
to make by adding non-DNF stochastic regular expressions as a new type of
``closed'' atom. When converted into a DNF regular expression, if $\Regex$ is
annotated as ``closed'', then $\ToDNFRegexOf{\Regex} = \AtomToDNFOf{\Regex}$.
Regular expressions that are not annotated as closed, are considered ``open.''

@@

Classical symmetric lenses satisfy the following equational laws.
\begin{equation}
  \tag{\PutRL}
  \frac{\PutRSymOf{(x,c)} = (y,c'')}{\PutLSymOf{(y,c'''')} = (x,c')}
\end{equation}
\begin{equation}
  \tag{\PutLR}
  \frac{\PutLSymOf{(y,c)} = (x,c')}{\PutRSymOf{(x,c'')} = (y,c')}
\end{equation}
